Range Analysis of Whole Programs

### 2012-02-23 doc/Manuscript/paper.tex
Added new references to the introduction.

% Our contribution.
In this paper we provide a complete description of a range analysis algorithm,
and show extensive experimental data that justifies our engineering choices.
Our main algorithmic contribution on top of previous works is a three-phases
approach to handle comparisons between variables without resorting to any
exponential time technique.
The few publicly available implementations of range analyses that we are
aware of, such as those in Java Hotspot~\footnote{} and gcc~\footnote{} only
deal with comparisons between variables and constants.
Even the theoretical works, such as Su and Wagner's~\cite{Su05} or Gawlitza
{\em et al.}'s~\cite{Gawlitza09} suffer from this limitation.
This deficiency is one of the reasons explaining why none of these works has
made their way into industrial-strength compilers.
Two other insights allow our implementation to scale up to very large
programs.
Firstly, we use Bodik's Extended Static Single Assignment (e-SSA) form to
perform range analysis sparsely.
This program representation ensures that the interval associated with
a variable is constant along its entire live range, and allows us to compute
path-sensitive results.
Secondly, we process the strongly connected components that underlines our
constraint system in topological ordering.
It is well-know that this technique is essential to speedup constraint solving
algorithms~\cite[Sec 6.3]{Nielson99}; however, we show that a careful
propagation of information along strong components not only gives us speed, but
also improves the precision of our results.

@@

% Our experimental numbers.
We have implemented our algorithm in the LLVM compiler~\cite{Lattner04}, and
have used it to analyze a test suite with 1.4 million lines of code.
In Section~\ref{sec:exp} we provide empirical evidence showing that our
implementation is fast: it analyzes the gcc source code in XX seconds, for
instance.
It is also precise: our results are similar to Stephenson {\em et al}'s,
even though, contrary to them, we do not assume that the program is correct.
Furthermore, we have been able to find tight bounds to many of the examples
used by Costan {\em et al.}~\cite{Costan05} and Lakhdar
{\em et al.}~\cite{Lakhdar11}, even though they use much more costly methods.

@@

\begin{figure}[t!]
\begin{small}
\begin{eqnarray*}
\begin{array}{r@{\hspace{0.5cm}}c}
Y = [l, u]
&
e(Y) = [l, u]
\\
\\
Y = \phi (X_1, X_2)
&
\inferrule{I[X_1]=[l_1, u_1] \\ I[X_2]=[l_2, u_2]}
{e(Y) = [\mbox{min}(l_1, l_2), \mbox{max}(u_1, u_2)]}
\\
\\
Y = X_1 + X_2
&
\inferrule{I[X_1]=[l_1, u_1] \\ I[X_2]=[l_2, u_2]}
{e(Y) = [l_1 + l_2, u_1 + u_2]}
\\
\\
Y = X_1 \times X_2
&
\inferrule{I[X_1]=[l_1, u_1] \\ I[X_2]=[l_2, u_2] \\ L = \{l_1l_2, l_1u_2, u_1l_2, u_1u_2\}}
{e(Y) = [\mbox{min}(L), \mbox{max}(L)]}
\\
\\
Y = aX + b
&
\inferrule{I[X]=[l, u] \\ k_l = al + b \\ k_u = au + b}
{e(Y) = [\mbox{min}(k_l, k_u), \mbox{max}(k_l, k_u)]}
\\
\\
Y = X \sqcap [l', u']
&
\inferrule{I[X]=[l, u]}
{e(Y) \leftarrow [\mbox{max}(l, l'), \mbox{min}(u, u')]}
\end{array}
\end{eqnarray*}
\caption{\label{fig:eval_function}
The constraint system that produces instances of the range analysis problem.}
\end{small}
\end{figure}

@@

\bibliographystyle{plain}
\bibliography{bibtex/references}

>>>>

% Our contribution.
In this paper we provide a complete description of a range analysis algorithm,
and show extensive experimental data that justifies our engineering choices.
Our first algorithmic contribution on top of previous works is a three-phases
approach to handle comparisons between variables without resorting to any
exponential time technique.
The few publicly available implementations of range analyses that we are
aware of, such as those in FLEX~\footnote{The MIT's FLEX/Harpoon compiler
provides an implementation of Stephenson's algorithm~\cite{Stephenson00}, and
is available at \texttt{http://flex.cscott.net/Harpoon/}.} and
gcc~\footnote{Gcc's VRP pass implements a variant of Patterson's
algorithm~\cite{Patterson95}} only deal with comparisons between variables and
constants.
Even theoretical works, such as Su and Wagner's~\cite{Su05} or Gawlitza
{\em et al.}'s~\cite{Gawlitza09} suffer from this limitation.
This deficiency is one of the reasons explaining why none of these works has
made their way into industrial-strength compilers.
Two other insights allow our implementation to scale up to very large
programs.
We use Bodik's Extended Static Single Assignment (e-SSA) form~\cite{Bodik00} to
perform range analysis sparsely.
This program representation ensures that the interval associated with
a variable is constant along its entire live range, and allows us to compute
path-sensitive results.
Finally, we process the strongly connected components that underline our
constraint system in topological ordering.
It is well-know that this technique is essential to speedup constraint solving
algorithms~\cite[Sec 6.3]{Nielson99}; however, we show that a careful
propagation of information along strong components not only gives us speed, but
also improves the precision of our results.

@@

% Our experimental numbers.
We have implemented our algorithm in the LLVM compiler~\cite{Lattner04}, and
have used it to analyze a test suite with 1.4 million lines of code.
In Section~\ref{sec:exp} we provide empirical evidence showing that our
implementation is fast: it analyzes the gcc source code in XX seconds, for
instance.
It is also precise: our results are similar to Stephenson
{\em et al}'s~\cite{Stephenson00}, even though, contrary to them, we do not
assume that the program is correct.
Furthermore, we have been able to find tight bounds to many of the examples
used by Costan {\em et al.}~\cite{Costan05} and Lakhdar
{\em et al.}~\cite{Lakhdar11}, who rely on much more costly methods.

@@

\begin{figure}[hbt]
\begin{small}
\begin{eqnarray*}
\begin{array}{r@{\hspace{0.5cm}}c}
Y = [l, u]
&
e(Y) = [l, u]
\\
\\
Y = \phi (X_1, X_2)
&
\inferrule{I[X_1]=[l_1, u_1] \\ I[X_2]=[l_2, u_2]}
{e(Y) = [\mbox{min}(l_1, l_2), \mbox{max}(u_1, u_2)]}
\\
\\
Y = X_1 + X_2
&
\inferrule{I[X_1]=[l_1, u_1] \\ I[X_2]=[l_2, u_2]}
{e(Y) = [l_1 + l_2, u_1 + u_2]}
\\
\\
Y = X_1 \times X_2
&
\inferrule{I[X_1]=[l_1, u_1] \\ I[X_2]=[l_2, u_2] \\ L = \{l_1l_2, l_1u_2, u_1l_2, u_1u_2\}}
{e(Y) = [\mbox{min}(L), \mbox{max}(L)]}
\\
\\
Y = aX + b
&
\inferrule{I[X]=[l, u] \\ k_l = al + b \\ k_u = au + b}
{e(Y) = [\mbox{min}(k_l, k_u), \mbox{max}(k_l, k_u)]}
\\
\\
Y = X \sqcap [l', u']
&
\inferrule{I[X]=[l, u]}
{e(Y) \leftarrow [\mbox{max}(l, l'), \mbox{min}(u, u')]}
\end{array}
\end{eqnarray*}
\caption{\label{fig:eval_function}
The constraint system that produces instances of the range analysis problem.}
\end{small}
\end{figure}

@@

\section{Experiments}
\label{sec:exp}

\bibliographystyle{plain}
\bibliography{references}
